Statistically-estimated tree composition for the northeastern United States \\ at the time of Euro-American settlement

### 2015-09-05 arxiv/paperArxiv.tex
added arxiv and plos versions of tex

\batchmode
\makeatletter
\def\input@path{{/accounts/gen/vis/paciorek/research/jmac/composition_paper//}}
\makeatother
\documentclass[12pt]{article}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

@@

@@

\makeatother
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

\title{Statistically-estimated tree composition for the northeastern United
States \\
at the time of European settlement}
\maketitle
\begin{quote}
Christopher J. Paciorek 1{*}, Simon J. Goring 2\&, Andrew Thurman
3\&, Charles V. Cogbill 4, \\
John W. Williams 3,5, David J. Mladenoff 6, Jody A. Peters 7, Jun
Zhu 8, \\
and Jason S. McLachlan 7 \\
{[}author order to be discussed{]}\\
{[}please tell me if you do/don't want your middle initial{]}

{[}let me know if there are any issues with how I've listed your affiliation{]}

1 Department of Statistics, University of California, Berkeley, California,
USA

2 Department of Geography, University of Wisconsin, Madison, Wisconsin,
USA

3 Department of Statistics, University of Iowa, Iowa City, Iowa, USA

4 Harvard Forest, Harvard University, Petersham, Massachusetts, USA

5 Center for Climatic Research, University of Wisconsin, Madison,
Wisconsin, USA

6 Department of Forestry and Wildlife Ecology, University of Wisconsin,
Madison, Wisconsin, USA

7 Department of Biological Sciences, University of Notre Dame, Notre
Dame, Indiana, USA

8 Department of Statistics, University of Wisconsin, Madison, Wisconsin,
USA

{*} Corresponding author

E-mail paciorek@stat.berkeley.edu (CJP)

\& These authors contributed equally to this work.

Short title: Settlement-era northeast U.S. tree composition
\end{quote}

\begin{abstract}
We present a data product of the estimated composition of tree taxa
at the time of European settlement of the northeastern United States
and the statistical methodology used to produce the product. Composition
is defined as the proportion of stems larger than approximately 20
cm diameter at breast height in 22 taxonomic groupings, generally
at the genus level. The data come from settlement survey records that
provide raw data that are transcribed and then aggregated spatially,
giving count data. The domain is divided into two regions, eastern
(Maine to Ohio) and western (Indiana to Minnesota). Public Land Survey
point data in the western region are aggregated to a regular 8 km
grid, while data in the eastern region, from Town Proprietor Surveys,
are aggregated at the township level in irregularly-shaped local administrative
units. The product is based on a Bayesian statistical model fit to
the count data that estimates composition on a regular 8 km grid.
The statistical model allows us to estimate composition at locations
with no data and to smooth over noise caused by limited counts in
locations with data. Critically, it also allows us to quantify uncertainty
in our composition estimates. We expect this data product to be useful
for understanding the state of vegetation in the northeastern United
States prior to large-scale European settlement. In addition to specific
regional questions, the data product can also serve as a baseline
against which to investigate how forests and ecosystems change after
intensive settlement.
\end{abstract}

@@

Historical datasets provide critical context to understand forest
ecology. They allow researchers to define `baseline' conditions for
conservation management, to understand ecosystem processes at decadal
and centennial scales, and, particularly in regions with widespread
land use change, to understand the extent to which forests after conversion
and regeneration differ from the original forest cover.

Euro-American settlement and subsequent land use change occurred in
a time-transient fashion across North America, and land surveys that
provide vegetation information followed a similar pattern. Early surveys
(from 1620 until 1825) in the northeastern United States provide areally-aggregated
data at the township level \citep{cogbill2002forests,thompson2013four};
we refer to these as the Town Proprietor Survey (TPS). Later surveys,
after the establishment of the U.S. Public Land Survey System (PLS)
provide point-level data along a regular grid, with one mile spacing
during the period 1832 to 1907 \citep{bourdo1956review,schulte2001original,goring2015composition}.
Survey instructions during the PLS varied through time and by point
type. Accounting for this requires the application of spatially-varying
correction factors \citep{cogbill2015corrections,goring2015composition}
to accurately assess stem density, basal area and biomass from the
early settlement records, but the impact on composition estimates
is limited. Both types of surveys provide composition of major tree
species but using sometimes inconsistent common names.

Logging, agriculture, and land abandonment have left an indelible
mark on forests in the northeastern United States \citep{foster1998land,rhemtulla2009legacies,thompson2013four,goring2015composition}.
However most studies have assessed these effects in individual states
or smaller domains \citep{friedman2005regional,rhemtulla2009historical}
and with varying spatial resolution, from townships (36 square miles)
to forest zones of hundreds or thousands of square miles. \citet{goring2015composition}
provide a dataset for the upper Midwest that is resolved to an 8 x
8 km grid size, providing broad spatial coverage at a spatial scale
that can be compared to the modern forests using Forest Inventory
and Analysis products \citep{gray2012forest}. Combined with additional,
coarsely-sampled PLS data from Illinois and Indiana and with the TPS
data, this gives us raw data for much of the northeastern United States.
However, there are several limitations of using the raw data that
can be alleviated by the use of a statistical model to develop a statistically-estimated
data product. First, the PLS and TPS data provide only estimates of
within-cell variance that do not account for information from nearby
locations. Second, the available digitized data from Illinois and
Indiana represent a small fraction of those states and missing townships
are common in the TPS data. Our statistical model allows us to provide
a spatially-complete data product of settlement-era tree composition
with uncertainty across the northeastern U.S.

@@

The raw data were obtained from survey records collated from across
the northeastern U.S. by a number of researchers. For the states of
Minnesota, Wisconsin, Illinois, Indiana, and Michigan (the western
subdomain), data are available at PLS survey point locations and have
been aggregated to a regular 8 km grid in the Albers projection. (Note
that for Indiana and Illinois, at the moment trees are associated
with township centroids and then assigned to 8 km grid cells based
on the centroid but in the near future we will have point locations
available for each tree.) For the states of Ohio, Pennsylvania, New
Jersey, New York and the six New England states (the eastern subdomain),
data are aggregated at the township level. There are also data from
a single township in Quebec and a single township in northern Delaware.
Data are essentially complete in Minnesota, Wisconsin and Michigan,
but data in Illinois and Indiana represent a sample of the full set
of grid cells, with survey record transcription ongoing. Data for
the remaining states are available for a subset of the full set of
townships covering the domain. Fig. \ref{fig:Spatial-domain} shows
the domain, indicating the grid cells and townships with data.

\begin{figure}
\label{fig:domain}\includegraphics[scale=1.3]{0_accounts_gen_vis_paciorek_research_jmac_composition_paper_fig1.pdf}

\caption{Spatial domain of the northeastern United States, with locations with
data shown in gray. Locations are grid cells in western portion and
townships in eastern portion. In addition to locations without data
being indicated in white, grid cells completely covered in water are
white (e.g., a few locations in the northwestern portion of the domain
in the states of Minnesota and Wisconsin).\label{fig:Spatial-domain}}
\end{figure}

Note that surveys occurred over a period of more than 200 years as
European colonists (before U.S. independence) and the United States
settled what is now the northeastern United States. Our estimates
are for the period of settlement represented by the survey data and
therefore are time-transgressive; they do not represent any single
point in time across the domain, but rather the state of the landscape
at the time just before settlement occurred \citep{Whit:1996,Cogb:etal:2002}.

Extensive details on the upper Midwest (Minnesota, Wisconsin, Michigan)
data and processing steps are available in \cite{Gori:etal:2015}.
The lower Midwest (Illinois, Indiana) data were purchased from the
Indiana State Archives (Indiana) and Hubtack Document Resources (hubtack.com;
Illinois) and processed using similar steps as for the upper Midwest
data. The TPS data were compiled by C.V. Cogbill from a myriad of
archival sources representing land division surveys conducted in connection
with local settlement. The complete raw vegetation data are available
through the NIS Data Portal (accession number and version to be inserted
when available). The aggregation into taxonomic groups is primarily
at the genus level, but is at the species level in some cases of monospecific
genera. We model the following 22 taxa plus an ``other hardwood''
category: Atlantic white cedar (\emph{Chamaecyparis thyoides}), Ash
(\emph{Fraxinus spp.}), Basswood (\emph{Tilia americana}), Beech (\emph{Fagus
grandifolia)}, Birch (\emph{Betula spp.}), Black gum/sweet gum (\emph{Nyssa
sylvatica} and \emph{Liquidambar styraciflua}), Cedar/juniper (\emph{Juniperus
virginiana} and \emph{Thuja occidentalis}), Cherry (\emph{Prunus spp.}),
Chestnut (\emph{Castanea dentata}), Dogwood (\emph{Cornus spp.}),
Elm (\emph{Ulmus spp.}), Fir (\emph{Abies spp.}), Hemlock (\emph{Tsuga
canadensis}), Hickory (\emph{Carya spp.}), Ironwood (\emph{Carpinus
caroliniana} and \emph{Ostrya virginiana}), Maple (\emph{Acer spp.}),
Oak (\emph{Quercus spp.}), Pine (\emph{Pinus spp.}), Poplar/tulip
poplar (\emph{Populus spp.}~and \emph{Liriodendron tulipifera}),
Spruce (\emph{Picea spp.}), Tamarack (\emph{Larix laricina}), and
Walnut (\emph{Juglans nigra}). Note that in several cases (black gum/sweet
gum, ironwood, poplar/tulip poplar, cedar/juniper), because of ambiguity
in the common tree names used by surveyors, a group represents trees
from different families and even orders. For the western subdomain
we do not fit statistical models for Atlantic white cedar and chestnut
as these have 0 and 7 trees present, respectively. The taxa grouped
into the other hardwood category are those for which fewer than roughly
2000 trees were present in the dataset; however we include Atlantic
white cedar explicitly despite it only having 336 trees in the dataset.
Although surveyors avoided using small trees, there was no consistent
lower diameter limit. The data, however, roughly represent trees greater
than 8 inches (\textasciitilde{}20 cm) diameter at breast height (dbh),
but with some trees as small as 3 inches dbh. In Indiana and Illinois
there are also 402 trees and 29 trees, respectively, that are between
1 and 4 inches dbh, but these will be excluded in the next version
of the product {[}waiting on Simon to see if he stripped them out{]}.

There are approximately 860,000 trees from the western subdomain and
420,000 trees from the eastern subdomain. In the western subdomain,
oak is the most common taxon and pine the second most common, while
in the eastern subdomain oak is the most common and beech the second
most common.

Our domain is a rectangle covering all of the states, in the Albers
projection (NAD 1983 Albers Great Lakes and St. Lawrence), with the
rectangle split into 8 km cells, arranged in a 296 by 180 grid of
cells, with the centroid of the cell in the southwest corner located
at (-71000, 58000). For the modeling of the western subdomain we use
the western-most 146 by 180 grid of cells. For the modeling of the
eastern subdomain we use the eastern-most 180 by 180 grid of cells
and then omit 23 cells in the north and 17 cells in the south outside
the states containing data.

@@

We fit a Bayesian statistical model to the data, with two primary
goals:
\begin{enumerate}
\item To estimate composition on a regular grid across the entire domain,
filling gaps where no data are available, and
\item To quantify uncertainty in composition at all locations. Even in grid
cells and townships with data, we wish to quantify uncertainty because
the empirical proportions represent estimates of the true proportions
that could be calculated using full population of all the trees in
an areal region.
\end{enumerate}
The result of fitting the Bayesian model via Markov chain Monte Carlo
(MCMC) is a set of representative samples from the posterior distribution
for the composition in the 23 taxonomic groupings at each of the grid
cells. These samples are the data product (described further in the
Data Product section) and can then be used in subsequent analyses.
The mean and standard deviation of the samples for each pair of cell
and taxon represent our best estimate (i.e., prediction) of composition
and a Bayesian ``standard error'' quantifying the uncertainty in
the estimate.

@@

The statistical model treats the observations as coming from a multinomial
distribution with a (latent) vector of proportions for each grid cell,
\[
y_{i}\sim\mbox{Multi}(n_{i},\theta(s_{i})),
\]
where $y_{i}$ is the vector of counts for the $P$ taxa at the $i$th
cell, $n_{i}$ is the number of trees counted in the cell, and $\theta(s_{i})$
is the vector of unknown proportions for those taxa at that cell.
Note that we use a standard multinomial distribution without overdispersion
as the set of trees in the dataset is roughly uniformly sampled across
the cells or townships \citep{Gori:etal:2015}.

@@

The critical component of the statistical model is the representation
of $\alpha_{p}(s)$ as a spatial process. This process is a prior
structure that serves to smooth across noise in the observations and
allows for interpolation to locations with no data. Apart from the
sum to one constraints, the taxa are considered to be independent
in the prior. We did not want to impose any structure that ties the
different taxa together, as any correlation will likely vary across
space.

In the next section, we consider two spatial models to define the
structure of the $\alpha_{p}(s)$ processes, a standard conditional
autoregressive model \citep{Bane:etal:2003} and a Gaussian Markov
random field (MRF) approximation to a Gaussian process with Matern
covariance \citep{Lind:etal:2011}.

@@

MRF models work directly with the precision matrix of the values of
the spatial process, so calculation of the prior density of $\alpha_{p}$
is computationally simple \citep{Rue:Held:2005}, but in situations
where the likelihood is not normal, it can be difficult to set up
effective MCMC algorithms that are able to move in the high-dimensional
space of $\alpha_{p}$. The latent variable representation of Section
\ref{sub:Model-for-township} helps to alleviate this problem.

@@

Gaussian processes (GP) are also standard models for spatial processes \citep{Bane:etal:2003}.
GP models are computationally challenging for large datasets because
of manipulations involving large covariance matrices. Given this,
\cite{Lind:etal:2011} proposed a new framework for using Gaussian
MRFs (GMRFs) as approximations to GPs, based on the use of stochastic
partial differential equations (SPDE).

>>>>

\documentclass[12pt]{article}
\usepackage[]{graphicx}
\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)

%\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
%\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
%\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
%\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
%\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
%\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
%\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
%\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
%\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
%\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%

%\usepackage{framed}
%\makeatletter
%\newenvironment{kframe}{%
% \def\at@end@of@kframe{}%
% \ifinner\ifhmode%
%  \def\at@end@of@kframe{\end{minipage}}%
%  \begin{minipage}{\columnwidth}%
% \fi\fi%
% \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
% \colorbox{shadecolor}{##1}\hskip-\fboxsep
%     % There is no \\@totalrightmargin, so:
%     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
% \MakeFramed {\advance\hsize-\width
%   \@totalleftmargin\z@ \linewidth\hsize
%   \@setminipage}}%
% {\par\unskip\endMakeFramed%
% \at@end@of@kframe}
%\makeatother

%\definecolor{shadecolor}{rgb}{.97, .97, .97}
%\definecolor{messagecolor}{rgb}{0, 0, 0}
%\definecolor{warningcolor}{rgb}{1, 0, 1}
%\definecolor{errorcolor}{rgb}{1, 0, 0}
%\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Textclass specific LaTeX commands.
\newcommand{\lyxaddress}[1]{
\par {\raggedright #1
\vspace{1.4em}
\noindent\par}
}

@@

\makeatother

\begin{document}

\title{Statistically-estimated tree composition for the northeastern United
States \\
at the time of Euro-American settlement}

\author{Christopher J. Paciorek\textsuperscript{1,{*}}, Simon J. Goring\textsuperscript{2,\&},
Andrew L. Thurman\textsuperscript{3,\&}, \\
Charles V. Cogbill\textsuperscript{4}, John W. Williams\textsuperscript{2,3},
David J. Mladenoff\textsuperscript{6}, \\
Jody A. Peters\textsuperscript{7}, Jun Zhu\textsuperscript{8}, and
Jason S. McLachlan\textsuperscript{7}}

\maketitle

\lyxaddress{\textsuperscript{1}Department of Statistics, University of California, Berkeley, California, USA}

\lyxaddress{\textsuperscript{2}Department of Geography, University of Wisconsin, Madison, Wisconsin, USA}

\lyxaddress{\textsuperscript{3}Department of Statistics, University of Iowa, Iowa City, Iowa, USA}

\lyxaddress{\textsuperscript{4}Harvard Forest, Harvard University, Petersham, Massachusetts, USA}

\lyxaddress{\textsuperscript{5}Center for Climatic Research, University of Wisconsin,
Madison, Wisconsin, USA}

\lyxaddress{\textsuperscript{6}Department of Forestry and Wildlife Ecology,
University of Wisconsin, Madison, Wisconsin, USA}

\lyxaddress{\textsuperscript{7}Department of Biological Sciences, University
of Notre Dame, Notre Dame, Indiana, USA}

\lyxaddress{\textsuperscript{8}Department of Statistics, University of Wisconsin,
Madison, Wisconsin, USA}

\lyxaddress{\textsuperscript{{*}}Corresponding author; E-mail: paciorek@stat.berkeley.edu}

\lyxaddress{\textsuperscript{\&}These authors contributed equally to this work.
\newpage}

\begin{abstract}
We present a gridded 8 km-resolution data product of the estimated
composition of tree taxa at the time of Euro-American settlement of
the northeastern United States and the statistical methodology used
to produce the product from trees recorded by land surveyors. Composition
is defined as the proportion of stems larger than approximately 20
cm diameter at breast height for 22 tree taxa, generally at the genus
level. The data come from settlement-era public survey records that
are transcribed and then aggregated spatially, giving count data.
The domain is divided into two regions, eastern (Maine to Ohio) and
midwestern (Indiana to Minnesota). Public Land Survey point data in
the midwestern region (ca. 0.8-km resolution) are aggregated to a
regular 8 km grid, while data in the eastern region, from Town Proprietor
Surveys, are aggregated at the township level in irregularly-shaped
local administrative units. The product is based on a Bayesian statistical
model fit to the count data that estimates composition on a regular
8 km grid across the entire domain. The statistical model is designed
to handle data from both the regular grid and the irregularly-sized
townships and allows us to estimate composition at locations with
no data and to smooth over noise caused by limited counts in locations
with data. Critically, the model also allows us to quantify uncertainty
in our composition estimates, making the product suitable for applications
employing data assimilation. We expect this data product to be useful
for understanding the state of vegetation in the northeastern United
States prior to large-scale Euro-American settlement. In addition
to specific regional questions, the data product can also serve as
a baseline against which to investigate how forests and ecosystems
change after intensive settlement. The data product is being made
available at the NIS data portal as version 0.4 under the CC BY license.

Keywords: biogeography, species composition, old-growth forests, spatial
modeling, Bayesian statistical model, vegetation mapping
\end{abstract}

@@

Historical datasets provide critical context to understand forest
ecology. They allow researchers to define `baseline' conditions for
conservation management, to understand ecosystem processes at decadal
and centennial scales, to track forest responses to shifting climates,
and, particularly in regions with widespread land use change, to understand
the extent to which forests after conversion and regeneration differ
from the original forest cover.

Euro-American settlement and subsequent land use change occurred in
a time-transient fashion across North America and were accompanied
by land surveys needed to demarcate land for land tenure and use.
Various systems were used by surveyors to locate legal boundary markers,
usually by recording and marking trees adjacent to survey markers.
These data provide vegetation information that can be mapped and used
quantitatively to represent the period of settlement. Early surveys
(from 1620 until 1825) in the northeastern United States provide spatially-aggregated
data at the township level \citep{Cogb:etal:2002,thompson2013four},
with typical township size on the order of 200 $\mbox{km}^{2}$ and
no information about the locations of individual trees; we refer to
these as the Town Proprietor Survey (TPS). Later surveys after the
establishment of the U.S. Public Land Survey System (PLS) by the General
Land Office (GLO) provide point-level data along a regular grid, with
one-half mile (800 m) spacing, for Ohio and westward during the period
1785 to 1907 \citep{bourdo1956review,pattison1957beginnings,schulte2001original,goring2015composition}.
At each point 2-4 trees were identified, and the common name, diameter
at breast height, and distance and bearing from the point were recorded.
Survey instructions during the PLS varied through time and by point
type. Accounting for this variation requires data screening to maximize
consistency among points and the application of spatially-varying
correction factors \citep{cogbill2015corrections,goring2015composition}
to accurately assess tree stem density, basal area and biomass from
the early settlement records, but the impact on composition estimates
is limited \citep{liu2011broadscale}. Surveyors sometimes used ambiguous
common names, which requires matching to scientific names and standardization
\citep{mladenoff2002narrowing,goring2015composition}.

Logging, agriculture, and land abandonment have left an indelible
mark on forests in the northeastern United States \citep{foster1998land,rhemtulla2009legacies,thompson2013four,goring2015composition}.
However most studies have assessed these effects in individual states
or smaller domains \citep{friedman2005regional,rhemtulla2009historical}
and with various spatial resolutions, from townships (36 square miles)
to forest zones of hundreds or thousands of square miles. \citet{goring2015composition}
provide a new dataset of forest composition, biomass, and stem density
based on PLS data for the upper Midwest that is resolved to an 8 km
by 8 km grid cell scale, providing broad spatial coverage at a spatial
scale that can be compared to modern forests using Forest Inventory
and Analysis products \citep{gray2012forest}. Combined with additional,
coarsely-sampled PLS data from Illinois and Indiana, newly-digitized
data from southern Michigan, and with the TPS data, this gives us
raw data for much of the northeastern United States. However, there
are several limitations of using the raw data that can be alleviated
by the use of a statistical model to develop a statistically-estimated
data product. First, the PLS and TPS data only provide estimates of
within-cell variance that do not account for information from nearby
locations. Second, there are data gaps: the available digitized data
from Illinois and Indiana represent a small fraction of those states
and missing townships are common in the TPS data. Third, the TPS and
PLS data have fundamentally different sampling design and spatial
resolution. Our statistical model allows us to provide a spatially-complete
data product of settlement-era tree composition for a common 8 km
grid with uncertainty across the northeastern U.S.

@@

The raw data were obtained from land division survey records collated
and digitized from across the northeastern U.S. by a number of researchers
(Fig. \ref{fig:Spatial-domain}). For the states of Minnesota, Wisconsin,
Illinois, Indiana, and Michigan (the midwestern subdomain), digitized
data are available at PLS survey point locations and have been aggregated
to a regular 8 km grid in the Albers projection. (Note that for Indiana
and Illinois, at the moment trees are associated with township centroids
and then assigned to 8 km grid cells based on the centroid but in
the near future we will have point locations available for each tree.)
For the states of Ohio, Pennsylvania, New Jersey, New York and the
six New England states (the eastern subdomain), data are aggregated
at the township level. There are also data from a single township
in Quebec and a single township in northern Delaware. Digitization
of PLS data in Minnesota, Wisconsin and Michigan is essentially complete,
with PLS data for nearly all 8 km grid cells, but data in Illinois
and Indiana represent a sample of the full set of grid cells, with
survey record transcription ongoing. Data for the eastern states are
available for a subset of the full set of townships covering the domain;
the TPS data for some townships were lost, incomplete, or have not
been located \citep{Cogb:etal:2002}.

\begin{figure}
\label{fig:domain}\includegraphics[scale=1.3]{fig1}

\caption{Spatial domain of the northeastern United States, with locations with
data shown in gray. Locations are grid cells in midwestern portion
and townships in eastern portion. In addition to locations without
data being indicated in white, grid cells completely covered in water
are white (e.g., a few locations in the northwestern portion of the
domain in the states of Minnesota and Wisconsin).\label{fig:Spatial-domain}}
\end{figure}

Note that surveys occurred over a period of more than 200 years as
European colonists (before U.S. independence) and the United States
settled what is now the northeastern and midwestern United States.
Our estimates are for the period of settlement represented by the
survey data and therefore are time-transgressive; they do not represent
any single point in time across the domain, but rather the state of
the landscape at the time just prior to widespread Euro-American settlement
and land use \citep{Whit:1996,Cogb:etal:2002}. These forest composition
datasets do include the effects of Native American land use and early
Euro-American settlement activities \citep[e.g.,][]{Blac:etal:2006},
but it is likely that the imprint of this earlier land use is highly
concentrated rather than spatially extensive \citep{munoz2014defining}.

Extensive details on the upper Midwest (Minnesota, Wisconsin, Michigan)
data and processing steps are available in \cite{goring2015composition};
key elements include the use of only corner points, the use of only
the two closest trees at each corner point, spatially-varying correction
factors for sampling effort, and a standardized taxonomy table. The
lower Midwest (Illinois, Indiana) data were purchased from the Indiana
State Archives (Indiana) and Hubtack Document Resources (hubtack.com;
Illinois) and processed using similar steps as for the upper Midwest
data. Digitization of the Illinois and Indiana data is still underway,
so many grid cells contained no data at the time the statistical model
was fit. Note that the number of trees per grid cell varies depending
on the number of survey points in a cell, with an average of 124 trees
per cell. The gridded data at the 8 km resolution for the upper Midwest
are available through the NIS Data portal (accession number and version
to be inserted when available). The TPS data were compiled by C.V.
Cogbill from a myriad of archival sources representing land division
surveys conducted in connection with local settlement.

The aggregation into taxonomic groups is primarily at the genus level
but is at the species level in some cases of monospecific genera.
We model the following 22 taxa plus an ``other hardwood'' category:
Atlantic white cedar (\emph{Chamaecyparis thyoides}), Ash (\emph{Fraxinus
spp.}), Basswood (\emph{Tilia americana}), Beech (\emph{Fagus grandifolia)},
Birch (\emph{Betula spp.}), Black gum/sweet gum (\emph{Nyssa sylvatica}
and \emph{Liquidambar styraciflua}), Cedar/juniper (\emph{Juniperus
virginiana} and \emph{Thuja occidentalis}), Cherry (\emph{Prunus spp.}),
Chestnut (\emph{Castanea dentata}), Dogwood (\emph{Cornus spp.}),
Elm (\emph{Ulmus spp.}), Fir (\emph{Abies balsamea}), Hemlock (\emph{Tsuga
canadensis}), Hickory (\emph{Carya spp.}), Ironwood (\emph{Carpinus
caroliniana} and \emph{Ostrya virginiana}), Maple (\emph{Acer spp.}),
Oak (\emph{Quercus spp.}), Pine (\emph{Pinus spp.}), Poplar/tulip
poplar (\emph{Populus spp.}~and \emph{Liriodendron tulipifera}),
Spruce (\emph{Picea spp.}), Tamarack (\emph{Larix laricina}), and
Walnut (\emph{Juglans spp.}). Note that in several cases (black gum/sweet
gum, ironwood, poplar/tulip poplar, cedar/juniper), because of ambiguity
in the common tree names used by surveyors, a group represents trees
from different genera or even families. For the midwestern subdomain
we do not fit statistical models for Atlantic white cedar and chestnut
as these species have 0 and 7 trees present, respectively. The taxa
grouped into the other hardwood category are those for which fewer
than roughly 2000 trees were present in the dataset; however we include
Atlantic white cedar explicitly despite it only having 336 trees in
the dataset because of specific ecological interest in Atlantic white
cedar wetlands.

Diameters are only recorded in the PLS data. Although surveyors avoided
using small trees, there was no consistent lower diameter limit. The
PLS data generally represent trees greater than 8 inches (\textasciitilde{}20
cm) diameter at breast height (dbh), but with some trees as small
as 1 inch dbh (smaller trees were much more common in far northern
Minnesota). TPS data have no information about dbh, but the trees
were large enough to blaze and are presumed to be relatively large
trees useful for marking property boundaries.

There are approximately 860,000 trees in the midwestern subdomain
and 420,000 trees in the eastern subdomain. In the midwestern subdomain,
oak is the most common taxon and pine the second most common, while
in the eastern subdomain oak is the most common and beech the second
most common.

Our domain is a rectangle covering all of the states using a metric
Albers (Great Lakes and St. Lawrence) projection (PROJ4: EPSG:3175),
with the rectangle split into 8 km cells, arranged in a 296 by 180
grid of cells, with the centroid of the cell in the southwest corner
located at (-71000 m, 58000 m). For the midwestern subdomain we use
the western-most 146 by 180 grid of cells. For the eastern subdomain
we use the eastern-most 180 by 180 grid of cells and then omit 23
rows of cells in the north and 17 rows of cells in the south, as these
grid cells are outside of the states containing data.

@@

We fit a Bayesian statistical model to the data, with two primary
goals:
\begin{enumerate}
\item To estimate composition on a regular grid across the entire domain,
filling gaps where no data are available, and
\item To quantify uncertainty in composition at all locations. Even in grid
cells and townships with data, we wish to quantify uncertainty because
the empirical proportions represent estimates of the true proportions
that could be calculated using the full population of all the trees
in grid cell or township.
\end{enumerate}
The result of fitting the Bayesian model via Markov chain Monte Carlo
(MCMC) is a set of representative samples from the posterior distribution
for the composition in the 23 taxonomic groupings at each of the grid
cells. These samples are the data product (described further in the
Data Product section) and can be used in subsequent analyses. The
mean and standard deviation of the samples for each pair of cell and
taxon represent our best estimate (i.e., prediction) of composition
and a Bayesian ``standard error'' quantifying the uncertainty in
the estimate.

@@

The statistical model treats the observations as coming from a multinomial
distribution with a (latent) vector of proportions for each grid cell,
\[
y_{i}\sim\mbox{Multi}(n_{i},\theta(s_{i})),
\]
where $y_{i}$ is the vector of counts for the $P$ taxa at the $i$th
cell, $n_{i}$ is the number of trees counted in the cell, and $\theta(s_{i})$
is the vector of unknown proportions for those taxa at that cell.
Note that we use a standard multinomial distribution without overdispersion,
because the set of trees in the dataset is roughly uniformly sampled
across the cells or townships \citep{goring2015composition}.

@@

The critical component of the statistical model is the representation
of $\alpha_{p}(s)$ as a spatial process. This process is a prior
structure that serves to smooth across noise in the observations and
allows for interpolation to locations with no data. Apart from the
sum-to-one constraint, the taxa are considered to be independent in
the prior. We did not want to impose any structure that ties the different
taxa together, as any correlation will likely vary across space.

In the next section, we consider two spatial models to define the
structure of the $\alpha_{p}(s)$ processes, a standard conditional
autoregressive model \citep{Bane:etal:2004} and a Gaussian Markov
random field (MRF) approximation to a Gaussian process with Matern
covariance \citep{Lind:etal:2011}.

@@

MRF models work directly with the precision matrix of the values of
the spatial process, so calculation of the prior density of $\alpha_{p}$
is computationally simple \citep{Rue:Held:2005}, but in situations
where the likelihood is not normal, it can be difficult to set up
effective MCMC algorithms that are able to move in the high-dimensional
space of $\alpha_{p}$. The latent variable representation of Section
\ref{sub:Latent-Variable-Model} helps to alleviate this problem.
Next we describe two alternative spatial models that we considered;
in Section \ref{sec:Model-comparison}, we use cross-validation to
choose between these two representations.

@@

Gaussian processes (GP) are also standard models for spatial processes \citep{Bane:etal:2004}.
GP models are computationally challenging for large datasets because
of manipulations involving large covariance matrices. Given this,
\cite{Lind:etal:2011} proposed a new framework for using Gaussian
MRFs (GMRFs) as approximations to GPs, based on the use of stochastic
partial differential equations (SPDEs).

We consider Gaussian processes in the \matern class, using the following
parameterization of the \matern correlation function, 
\begin{equation}
R(d)=\frac{1}{\Gamma(\nu)2^{\nu-1}}\left(\frac{2\sqrt{\nu}d}{\rho}\right)^{\nu}\mathcal{K}_{\nu}\left(\frac{2\sqrt{\nu}d}{\rho}\right),\label{eq:Matern}
\end{equation}
where $d$ is Euclidean distance, $\rho$ is the spatial range parameter,
and $\mathcal{K}_{\nu}(\cdot)$ is the modified Bessel function of
the second kind, whose order is the smoothness (differentiability)
parameter, $\nu>0$. $\nu=0.5$ gives the exponential covariance.
For any pair of locations, $R(d)$ defines the correlation of the
process, (i.e., $\alpha_{p}(s)$ in our context), as a function of
the distance between the locations. Considering all pairs of locations,
this defines a correlation matrix for all locations of interest.
